A geometric framework for capturing high-resolution insights into conceptual knowledge and learning in classroom-like settings

### 2023-01-30 paper/main.tex
fixed initials in author contributions
rewrote topic modeling results paragraph

We asked participants in our study to complete brief multiple-choice quizzes before, between, and after watching two lecture videos from the Khan Academy \citep{Khan04} platform (Fig.~\ref{fig:experiment}). The first lecture video, entitled \textit{Four
Fundamental Forces}, discussed the four fundamental forces in physics: gravity,
strong and weak interactions, and electromagnetism. The second,
entitled \textit{Birth of Stars}, provided an overview of our current
understanding of how stars form.
%(See \textit{\nameref{subsec:experiment}} for more information on how these lectures were chosen.)
%Each multiple-choice quiz consisted of 13 questions that would enable us to test participants'
%knowledge about each individual lecture, as well as related content not
%specifically presented in either video ( see Tab.~\questions).
%
%We selected these lessons to be accessible to a broad audience (i.e., requiring minimal prerequisite knowledge to understand) and largely independent of each other (i.e., focused on sufficiently different material that understanding one did not require having seen the other), but conceptually related to each other (covering at least \textit{some} similar or overlapping content).
We selected these lessons to be (a) accessible
to a broad audience, i.e., requiring minimal prerequisite knowledge to understand;
(b) conceptually related to each other, i.e., covering at least \textit{some}
similar or overlapping content; and (c) largely independent of each other, i.e., focused
on sufficiently different material that understanding one did not require having seen
the other. The two videos we selected are introductory lectures that both
belong to Khan Academy's ``Cosmology and Astronomy'' course domain, but are taken from
different lecture series (``Scale of the Universe'' and ``Stars, Black Holes, and Galaxies''
for the first and second lectures, respectively).

%We asked participants in our study to answer questions from several multiple-choice
%quizzes and watch two lecture videos from the Khan Academy platform
%(Fig.~\ref{fig:experiment}). The first lecture video, entitled \textit{Four
%Fundamental Forces}, discussed the four fundamental forces in physics: gravity,
%strong and weak interactions, and electromagnetism. The second,
%entitled \textit{Birth of Stars}, provided an overview of our current
%understanding of how stars form. We selected these lessons to be (a) accessible
%to a broad audience, i.e., requiring minimal prerequisite knowledge to understand;
%(b) conceptually related to each other, i.e., covering at least \textit{some}
%similar or overlapping content; and (c) largely independent of each other, i.e., focused
%on sufficiently different material that understanding one did not require having seen
%the other. Thus, the two videos we selected are introductory, high-level lectures that both
%belong to Khan Academy's ``Cosmology and Astronomy'' course domain, but come from
%different lecture series (``Scale of the Universe'' and ``Stars, Black Holes, and Galaxies''
%for the first and second lectures, respectively).

@@

To study how participants' conceptual knowledge changed over the course of the experiment, 
we first sought to characterize the abstract concepts presented to them in each of the two 
lectures. We adapted an approach we developed in prior work \citep{HeusEtal21} to extract 
the latent themes from the lectures' contents using a topic model \citep{BleiEtal03}. Briefly, 
topic models take as input a collection of text documents and learn a set of ``topics'' (i.e., 
latent themes) from their contents. Once fit, a topic model can be used to transform arbitrary 
(potentially new) documents into sets of ``topic proportions,'' describing the weighted blend 
of learned topics reflected in their texts. We parsed automatically generated transcripts of 
the two lectures into overlapping sliding windows, and used these as documents to fit our 
model (Fig.~\ref{fig:sliding-windows}A; see \textit{\nameref{subsec:topic-modeling}}).
The ($k$=15) topics

%To construct a corpus of text documents for the two lectures, we parsed their transcripts into overlapping sliding windows (Fig.~\ref{fig:sliding-windows}A; see \textit{\nameref{subsec:topic-modeling}})

We trained a topic model \citep{BleiEtal03} using sliding windows of text from the two
lectures' transcripts (see \textit{\nameref{subsec:topic-modeling}}). We also used the same model (i.e., trained on the
videos' transcripts) to embed the text of each question in our pool. This
yielded, for each second of each video, and for each question, a single
15-dimensional topic vector---i.e., a coordinate in a text embedding space
(Fig.~\ref{fig:sliding-windows}). Intuitively, each dimension of the embedding
space corresponds to a ``theme'' or ``topic'' reflected in some part(s) of the
videos (Tab.~\topics), and the coordinates in embedding space denote the blend
of themes reflected by a particular excerpt of text (e.g., from part of a
video's transcript, from a question, etc.).

@@

Conceptualization: PCF, ACH, and JRM. Methodology: PCF, ACH, and JRM. Software:
PCF. Validation: PCF. Formal analysis: PCF. Resources: PCF, ACH, and JRM. Data
curation: PCF. Writing (original draft): JRM. Writing (review and editing):
PCF, ACH, and JRM. Visualization: PCH and JRM. Supervision: JRM. Project
administration: PCH. Funding acquisition: JRM.

>>>>

We asked participants in our study to complete brief multiple-choice quizzes before, between, and after watching two lecture videos from the Khan Academy \citep{Khan04} platform (Fig.~\ref{fig:experiment}). The first lecture video, entitled \textit{Four
Fundamental Forces}, discussed the four fundamental forces in physics: gravity,
strong and weak interactions, and electromagnetism. The second,
entitled \textit{Birth of Stars}, provided an overview of our current
understanding of how stars form.
We selected these lessons to be (a) accessible
to a broad audience, i.e., requiring minimal prerequisite knowledge to understand;
(b) conceptually related to each other, i.e., covering at least \textit{some}
similar or overlapping content; and (c) largely independent of each other, i.e., focused
on sufficiently different material that understanding one did not require having seen
the other. The two videos we selected are introductory lectures that both
belong to Khan Academy's ``Cosmology and Astronomy'' course domain, but are taken from
different lecture series (``Scale of the Universe'' and ``Stars, Black Holes, and Galaxies''
for the first and second lectures, respectively).

@@

To study how participants' conceptual knowledge changed over the course of the
experiment, we first sought to characterize the abstract concepts presented to
them in each of the two lectures. We adapted an approach we developed in prior
work \citep{HeusEtal21} to extract the latent themes from the lectures' contents
using a topic model \citep{BleiEtal03}. Briefly, topic models take as input a
collection of text documents and learn a set of ``topics'' (i.e., latent
themes) from their contents. Once fit, a topic model can be used to transform
arbitrary (potentially new) documents into sets of ``topic proportions,''
describing the weighted blend of learned topics reflected in their texts. We
parsed automatically generated transcripts of the two lectures into
overlapping sliding windows, which we treated as documents to fit our model
(Fig.~\ref{fig:sliding-windows}A; see \textit{\nameref{subsec:topic-modeling}}).
Transforming these windows with our model resulted in a number-of-windows by
number-of-topics (15) topic-proportions matrix, denoting the unique mixture
of broad themes from both lectures reflected in each window's content.
Intuitively, each window's ``topic vector'' (i.e., column of the
topic-proportions matrix) is analogous to a coordinate in a 15-dimensional
space (whose axes are topics learned by the model). Within this space, each lecture's series of
topic vectors (i.e., corresponding to sliding windows parsed from its
transcript) forms a \textit{trajectory} that captures how its conceptual
content unfolds over time (Fig.~\ref{fig:sliding-windows}A). We resampled
these trajectories to a resolution of 1 topic vector for each second of video.

@@

Conceptualization: PCF, ACH, and JRM. Methodology: PCF, ACH, and JRM. Software:
PCF. Validation: PCF. Formal analysis: PCF. Resources: PCF, ACH, and JRM. Data
curation: PCF. Writing (original draft): JRM. Writing (review and editing):
PCF, ACH, and JRM. Visualization: PCF and JRM. Supervision: JRM. Project
administration: PCF. Funding acquisition: JRM.
